\documentclass{beamer}

\usepackage{amsmath, amssymb, hyperref, graphics, wasysym, tikz}
\usetikzlibrary{cd}
\usepackage{mathpazo}

\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathbb{Q}}

\title{MAS439 Lecture 9 \\ Noetherian Rings}

\date{October 27th}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}{Noetherian Rings}

\begin{definition} We say that a ring $R$ is \emph{Noetherian} if it satisfies the \emph{Ascending Chain Condition}: any increasing chain of ideals eventually stabilizes.  That is, if
$$I_1\subset I_2\subset \dots\subset I_n\subset\dots \subset R$$
is a chain of ideals, then there exists some $N$ so that $I_n=I_n$ for all $n>N$
\end{definition}




\end{frame}
\begin{frame}{Why care about Noetherian rings?}

\begin{lemma}
A ring $R$ is Noetherian if and only if every ideal $I$ of $R$ is finitely generated
\end{lemma}


\end{frame}

\begin{frame}{Suppose that $R$ has every ideal finitely generated}

\begin{block}{How to show $R$ Noetherian?}
Given an ascending chain $I_1\subset I_2\subset I_3\subset\cdots \subset I_n\subset \cdots $, we must show the chain stabilizes.
\end{block}

Only thing we can do is use that some ideal $I$ is finitely generated.  Trying to use this on the $I_k$ doesn't help.  What other ideal could we use?






\end{frame}

\begin{frame}{A lemma we've already used...}

\begin{lemma}
Suppose $I_1\subset I_2\subset\dots\subset I_n\subset\dots$ ideals.  Then

$$I=\union_n I_n$$ is an ideal
\end{lemma}

\begin{proof}
\begin{itemize}
\item Nonempty as it contains $0$
\item Closed under addition: if $r,s\in I$, then $\exists n,m$ with $r\in I_n, s\in I_m$.  Let $k=\textrm{max}(n,m)$; then $r+s\in I_k\subset I$.
\item Closed under multiplication by $R$: if $x\in I, r\in I$, then $x\in I_n$ for some $n$, so $r\cdot x\in I_n\subset I$
\end{itemize}
\end{proof}
\end{frame}


\begin{frame}{Back to Noetherian rings}

Assume every ideal of $R$ is finitely generated, and $I_1\subset I_2\subset\dots \subset I_n\subset\dots $ a chain of ideal; need to show the ideal stabilizes.

\begin{block}{But $I=\union I_n$ is finitely generated}
Suppose that $I=(x_1,\dots, x_k)$.  
\end{block}

Then $x_i\in I$, so $x_i\in I_{n_i}$ for some $n_i$.  Taking $N=\textrm{max}(n_k)$, we see that $x_i\in I_N$.  Hence:

$$I=(x_i)\subset I_N\subset I$$
and so the chain stabilizes at $I_N$ to $I$.
\end{frame}

\begin{frame}{Noetherian $\implies$ every ideal finitely generated}

Now suppose that $R$ Noetherian, and $I\subset R$ an ideal.  We want to show that $I$ is finitely generated.  

\begin{block}{Where can we find a chain of ideals to apply the ACC to?}
Try to generate $I$ at random
\end{block}



\end{frame}

\end{document}


