\documentclass{amsart}[12pt]



\title{MAS439 Commutative Algebra \\ Problem set 2}
\author{Due Friday, November 1st}
\begin{document}
\maketitle
\section*{Question 1: A subring}
For $f$ \ polynomial, let $f^\prime$ be its derivative and $f^{\prime\prime}$ be its second derivative.  Define
$$S=\{f\in\mathbb{Z}[x] : f^\prime(1)=f^{\prime\prime}(1)=0\}$$
\subsection*{Part 1}
Prove that $S$ is a subring of $\mathbb{Z}[x]$.
\subsection*{Part 2} Prove that $S$ is finitely generated.

\section*{Question 2: Ideals?}
Let $R$ be a commutative ring.
\subsection*{Part 1} Let $N$ be the union of $\{0\}$ and all the nilpotent elements of $R$.  Prove that $N$ is an ideal of $R$.
\subsection*{Part 2} Let $D$ be the union of $\{0\}$ and all the zero divisors of $R$.  Give an example to show that $D$ need not be an ideal.

\section*{Question 3: Rings with four elements}
We already know three rings with four elements: $\mathbb{Z}/4, \mathbb{Z}/2\times\mathbb{Z}/2$ and $\mathbb{F}_4$, the field with four elements, which we constructed in lecture and in the notes as the quotient ring $\mathbb{F}_4=\mathbb{F}_2[x]/(x^2+x+1)$.

In addition to $x^2+x+1$, there are three more quadratic polynomials in $\mathbb{F}_2[x]$ (Recall that $\mathbb{F}_2$ is just what we write for $\mathbb{Z}/2$ when we want to emphasize it's a field):

$$f_1=x^2,\quad f_2=x^2+x,\quad f_3=x^2+1$$ 

\subsection*{Part one} Prove that each of the rings $R_i=\mathbb{F}_2[x]/(f_i)$ has four elements -- do this for all three at one time.

\subsection*{Part two} Write down the multiplication tables for each of the rings $R_i$, and for each one determine:
\begin{enumerate}
\item Is $R_i$ reduced?  
\item Is $R_i$ an integral domain?
\item Is $R_i$ a field?
\item Is $R_i$ isomorphic to one of three rings with four elements we already knew?
 \end{enumerate} 
\subsection*{Part three}
Finally, decide which of the $R_i$ are isomorphic to each other, and which are not? 



\end{document}
