\documentclass{amsart}[12pt]



\title{MAS439 Commutative Algebra \\ Problem set 3}
\author{Due Friday, November 22nd}
\begin{document}
\maketitle
\section{Ideals in product rings}
\subsection*{Ideals in products are products of ideals}
Prove that the ideals of $R\times S$ are \emph{precisely} the sets of the form
$$I\times J \{(x,y):x\in I, y\in J\} $$
where $I\subset R$ and $J\subset S$ are ideals.

\subsection*{Ideals of $\mathbb{Z}/2\times\mathbb{Z}/4$}
What are all the ideals of $\mathbb{Z}/2\times \mathbb{Z}/4$?  Which ones are radical?  Prime?  Maximal?  Hint: to explain and figure out this last one, it might be easier to look at the quotient ring $(\mathbb{Z}/2\times\mathbb{Z}/4)/I$.


\section{A nonfinitely generated ideal}
Later we will prove that if $k$ is a field, then every ideal in $k[x_1,\dots, x_n]$ is finitely generated.  Before we do so, it may be beneficial to have see an example of an ideal that \emph{isn't} finitely generated.

Let $R$ be the ring of continuous functions $f:\mathbb{R}\to \mathbb{R}$ with the usual pointwise addition and multiplication.  We say that a function $f\in R$ has bounded support if there is a real number $M>0$ such that

$$|x|>M\implies f(x)=0$$

\begin{enumerate}
  \item   Prove that the subset $I\subset R$ of functions with bounded support is an ideal of $R$.
\item Prove that $I$ is radical but not prime.
\item Prove that $I$ is not finitely generated.
\end{enumerate}

\section*{Question 1: Dimensions of $\mathbb{C}$-algebras}
Compute the dimensions of the following $\mathbb{C}$-algebras, that are all quotients of $\C[x,y]$ by ideals generated by a few elements.  Note that they might be infinite dimensional...  In each case give a basis for the algebra as a complex vector space.  Some explanation of your answer is required, but ironclad proofs are unnecessary.

\begin{enumerate}
\item $R_1=\mathbb{C}[x,y]/(x^3,xy,y^2)$
\item $R_2=\mathbb{C}[x,y]/(x^2y, x^3)$
\item $R_3=\mathbb{C}[x,y]/(x^2+y^2, x-y^3)$

\end{document}
